\Large \bf \Toolname: Cache Line Contention Detection for COTS Binaries

### 2014-12-17 HotOS'15/paper.tex
Working on the methodology section still. Kind of boring so far.

@@

% --------------------------------------
\subsection{Ownership Tracking}
\begin{figure}
\begin{center}
\epsfig{file=states.eps, height=30em}
\end{center}
\caption{\label{fig:states_of_shadow}States of individual shadow memory cells. Each cell tracks the watched and ownership %
status of a cache line. Yellow states are unwatched. Green states are watched. Red states represent contention that was missed %
due to races in the instrumentation. Transitions between states are based on (potentially stale) observed prior states. For %
example, the spurious acquire transition is possible because the decision to make the transition is based on a racy \texttt{CMP} %
against the $SampleId$ (\figref{inline_assembly}). If two threads simultaneously see that the cache line it watched but not %
owned, then both will execute the atomic \texttt{XCHG}, and only one thread will ``win". The ignore transition is induced, %
either because the sampling timeout expires, or more likely, because the report transition occurs.}
\end{figure}

\begin{figure}
\lstset{language=[x64]Assembler}
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
;; Racy watchpoint check.
CMP [ShadowAddr + Offs(SampleId)], 0
JZ ignore
;; Racy ownership check.
CMP [ShadowAddr + Offs(ThreadId)], CurrThreadId
JZ ignore
;; Take ownership, or transfer ownership
LOCK XCHG [ShadowAddr], OldState
CALL CheckContention(OldState,Addr,PC)
ignore:
\end{lstlisting}
\caption{\label{fig:inline_assembly}Instrumentation instructions injected for each shared memory access. The algorithm optimizes on %
instruction footprint by using a single \texttt{XCHG} instruction to perform multiple state transitions from %
\Figref{states_of_shadow}, and then figuring out what state we should be in based on the value of $OldState$.}
\end{figure}

@@

The limitation to 16 bits for $SampleId$s was a deliberate decision. We take advantage of the structure of ``canonical addresses" on the
x86-64 architecture, whereby the high-order 16 bits of a 64-bit address must have the same value as the 47th bit. In user space, all 16
high-order bits are 0. We represent the $ThreadId$ using the thread's ``base pointer'' (a pointer thread-local storage). Thus, the
\texttt{XCHG} in \Figref{inline_assembly} has the side-effect of temporarily disabling the watchpoint.

% TODO: describe why this is reasonable, what alternatives there are, and the complications with those alternatives (atomicity,
% re-watching, etc)

% TODO: Should we have done it where we read the value of shadow memory into a reg, then used that as the old state to pass on
% (plus the old value of the thread pointer)? This is valid, and might be better. But this has weird transitions.

% Essentially: the "right" solution using a CMPXCHG, but that requires a lot of spills/fills of regs  === heavyweight.

>>>>

% For algorithms
\usepackage{program}

@@

% --------------------------------------
\subsection{Ownership Tracking}
\begin{figure}
\begin{center}
\epsfig{file=states.eps, height=30em}
\end{center}
\caption{\label{fig:states_of_shadow}States of individual shadow memory cells. Shadow memory cells begin as unwatched. %
Then, when a cache line is selected for sampling, it becomes watched. This is represented by the presence of a $SampleId$. %
The first thread that tries to access the shadow memory of the cache line will become the cache line's owner, as represented %
by the presence of a $ThreadId$. If multiple threads attempt to simultaneously acquire ownership of the cache line, then %
contention will be detected and reported. If not, then execution will continue until a second thread accesses the cache line, %
thus triggering an ownership transfer.}
\end{figure}

\begin{figure}
\lstset{language=[x64]Assembler}
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
;; Racy watchpoint check.
CMP [Shadow(Addr) + Offs(SampleId)], 0
JZ ignore
;; Racy ownership check.
CMP [Shadow(Addr) + Offs(ThreadId)], MyThreadId
JZ ignore
;; Take ownership, or transfer ownership
CALL CheckContention(Addr,PC)
ignore:
\end{lstlisting}
\caption{\label{fig:inline_assembly}Instrumentation instructions injected for each shared memory access. The algorithm %
optimizes on instruction footprint by performing lightweight, racy checks up front as a way of avoiding more heavyweight %
contention detection. If the cache line appears watched, and if the race checks fail, then the $CheckContention$ function %
is invoked. This function handles the state transitions of \Figref{states_of_shadow} internally using an atomic compare and %
swap.}
\end{figure}

\begin{figure}
\begin{lstlisting}[language=C,basicstyle=\footnotesize\ttfamily]
void CheckContention(uintptr_t addr, PC pc) {
  // Figure out address of shadow memory,
  // given address of memory being accessed.
  WP *wp = (addr / 64) + gShadowBase;
  uint16_t sample_id = wp->sample_id;
  uint16_t thread_id = GetThreadId();
  if (!sample_id) return;

  // Assume unowned; try to take ownership. We
  // check the return value to figure out if
  // the prior state was unowned.
  WP state = CAS(wp, WP(sample_id,0),
                     WP(sample_id,thread_id));

  // Contention already detected, or
  // sampling disabled due to sample
  // period expiring.
  if (!state.sample_id) return;

  // Acquired ownership.
  if (!state.thread_id) {
    ReportOwner(sample_id,addr, pc);
    PauseThread(gConfigSleepTime);
  } else {
    wp->sample_id = 0;  // Unwatch.
    ReportContender(sample_id,addr, pc)
  }
}
\end{lstlisting}
\caption{\label{fig:algorithm}Algorithm for detecting contention on a cache line. Contention reports are updated as the threads %
are discovered, and finalized at the end of each sampling period before new samples are taken. The \texttt{PauseThread} function %
is notable. By passing a configurable sleep duration to the \texttt{PauseThread} function, we can turn on DataCollider-like detection, %
or by not pausing, turn on Dr. Memory-like detection. Not shown in this algorithm is the extent of information passed to the reports %
(memory operand sizes and types, call stacks).}
\end{figure}

@@

If a sampled memory location is accessed, then the instrumentation instructions of \Figref{inline_assembly} will detect the non-zero
$SampleId$, and call out to the $CheckContention$ function, but only if the current thread does not appear to own the cache line
already. The $CheckContention$ algorithm (\Figref{algorithm}) performs both the ``Acquire" and the ``Detect" state transitions of
\Figref{states_of_shadow} using a single atomic compare-and-swap operation (\texttt{CAS}).
